---
title: "Module QG3 - Association mapping: GWAS & sequencing data"
---

```{r echo = FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
```

**Instructors:** Joelle Mbatchou & Moeen Riaz

You will find on this website link to lecture slides and exercises.

The official SISG web page can be found [here](https://sisg.biosciences.gatech.edu/sisg-module-qg3/) – course evaluations and certificates of completion are available via this page.


## Session Format
The module has 10 sessions, each of 90 minutes. The standard format for a session is approximately:

* 45 minutes of lecture material
* 30 minutes of exercises for you to try
* 15 minute discussion of exercises, where the instructors will present possible solutions and Q&A

Each session builds on previous sessions so you will get maximum benefit by attending every session and in sequence.

The course schedule is below.


## Pre-requisites

We will be running exercises during the course using various tools, follow the instructions below to get setup on your machine.

**Prior to the module**, please install working and latest versions of:

* R (Version 4.1+):
  <a href="https://www.r-project.org/" target="_blank">https://www.r-project.org/</a>
* RStudio (Version 1.4+) :
  <a href="https://rstudio.com/products/rstudio/download/" target="_blank">https://rstudio.com/products/rstudio/download/</a>
* PLINK 1.9 :
  <a href="https://www.cog-genomics.org/plink2/" target="_blank">https://www.cog-genomics.org/plink2/</a>
* PLINK 2.0 :
  <a href="https://www.cog-genomics.org/plink/2.0/" target="_blank">https://www.cog-genomics.org/plink/2.0/</a>
* REGENIE :
  <a href="https://github.com/rgcgithub/regenie/releases/" target="_blank">https://github.com/rgcgithub/regenie/releases/</a>
* GCTA :
  <a href="https://yanglab.westlake.edu.cn/software/gcta/#Download" target="_blank">https://yanglab.westlake.edu.cn/software/gcta/#Download</a>
* GCTB :
  <a href="https://cnsgenomics.com/software/gctb/#Download" target="_blank">https://cnsgenomics.com/software/gctb/#Download</a>

All are free software. For those on Windows, we recommend WSL ([Windows Subsystem for Linux](https://learn.microsoft.com/en-us/windows/wsl/install)) to run the tools above if no Windows binary is available. Alternatively, you can also use [docker](https://docs.docker.com/desktop/setup/install/windows-install/) if docker images are available.

Please be sure to install some necessary R packages by running the following commands in a R terminal:
```{r install-pkgs, eval = FALSE}
install.packages("data.table")
install.packages("ggplot2")
install.packages("dplyr")
install.packages("qqman")
install.packages("SKAT")
install.packages("bigsnpr")
install.packages("hexbin")
install.packages("devtools")
install.packages("tidyverse")
install.packages("cowplot")
install.packages("BEDMatrix")
install.packages("glue")
devtools::install_github("yaowuliu/ACAT")
```

## Schedule

All times listed below for the schedule are Eastern Daylight Time (EDT, Atlanta time).

```{r mon-sched, echo = FALSE}
data.frame(
  Time = c("8:30am-10am", "10:30am-12pm", "1:30pm-3pm", "3:30pm-5pm"),
  Session = 1:4,
  Topic = c(
    "Case Control Association Testing & Association Testing with Quantitative Traits",
    "Introduction to the PLINK Software for GWAS & Population Structure Inference",
    "GWAS in Samples with Structure & Introduction to the REGENIE Software",
    "Rare Variant Analysis: Collapsing Tests, Kernel (Variance Component) Tests and Omnibus Tests"
    ),
  Lecture = c(
    '<a href="https://gatech.box.com/s/2xjedvhnv1j0jxczdrl9brw0peuza5ii" target="_blank">Slides</a>',
    '<a href="https://gatech.box.com/s/3ogiqqvtljgmrv1kp8ytv6m4yhyeqeoq" target="_blank">Slides</a>',
    '<a href="https://gatech.box.com/s/jt8f6977ji619pk64prfwhsvf7ecec45" target="_blank">Slides</a>',
    '<a href="https://gatech.box.com/s/soukkj3xew95gbbh30ft3ljyt19q94v5" target="_blank">Slides</a>'
  ),
  Exercises = c(
    '<a href="analysis/QG3_Association_Testing_practical.html">Practical</a>',
    '<a href="analysis/QG3_Plink_Population_Structure_practical.html">Practical</a>',
    '<a href="analysis/QG3_Relatedness_REGENIE_practical.html">Practical</a>',
    '<a href="analysis/QG3_RV_tests_practical.html">Practical</a>'
  ),
  Key = c("", "", "", "")
) %>%
kable(escape = FALSE, caption = "**Monday, June 8th**", align = c('l', 'c', 'l', 'c', 'c', 'c')) %>%
kable_styling(bootstrap_options = c("hover", "condensed", "striped"),
              full_width = FALSE,
              font_size = 14) %>%
column_spec(1, width = "12%") %>%
column_spec(2, width = "8%") %>%
column_spec(3, width = "35%")
```

```{r tue-sched, echo = FALSE}
data.frame(
  Time = c("8:30am-10am", "10:30am-12pm", "1:30pm-3pm", "3:30pm-5pm"),
  Session = 4 + 1:4,
  Topic = c(
    "Meta-analysis: GWAS meta-analysis with METAL, QC and best practices, Gene-based ExWAS with REMETA",
    "Meta-analysis with REMETA continue & Downstream analysis: Conditional analysis, fine mapping",
    "Downstream analysis continues: Coloc, Gene-based & Pathways Analysis",
    "Beyond traditional GWAS: Variance QTL, Interaction Testing, GWAX, Time-to-event, Multi-trait analysis"
    ),
  Lecture = c(
    '<a href="https://gatech.box.com/s/eaiwddxx1hs5dn3itz7pjjgjgjdfieg5" target="_blank">Slides</a>',
    '<a href="https://gatech.box.com/s/eaiwddxx1hs5dn3itz7pjjgjgjdfieg5" target="_blank">Slides</a>',
    "",
    '<a href="https://gatech.box.com/s/163v98wn6wrq1yib08cee3ck1wm1quje" target="_blank">Slides</a>'
  ),
  Exercises = c(
    '<a href="https://moeenriaz2.github.io/SISG26/index.html" target="_blank">Practical</a>',
    '<a href="https://moeenriaz2.github.io/SISG26/index.html" target="_blank">Practical</a>',
    '<a href="analysis/QG3_Downstream-Analyses_practical.html">Practical</a>',
    '<a href="analysis/QG3_Beyond_Standard_GWAS_practical.html">Practical</a>'
  ),
  Key = c("", "", "", "")
) %>%
kable(escape = FALSE, caption = "**Tuesday, June 9th**", align = c('l', 'c', 'l', 'c', 'c', 'c')) %>%
kable_styling(bootstrap_options = c("hover", "condensed", "striped"),
              full_width = FALSE,
              font_size = 14) %>%
column_spec(1, width = "12%") %>%
column_spec(2, width = "8%") %>%
column_spec(3, width = "35%")
```

```{r wed-sched, echo = FALSE}
data.frame(
  Time = c("8:30am-10am", "10:30am-12pm"),
  Session = 8 + 1:2,
  Topic = c(
    "Polygenic Risk Scores",
    "Emerging issues showcasing ongoing research"
    ),
  Lecture = c(
    '<a href="https://gatech.box.com/s/gr7jq894brykdq4s64nqd09b5guateel" target="_blank">Slides</a>',
    '<a href="https://gatech.box.com/s/jyw7gswr67n01zr1rt8b16xr8ptjdgaj" target="_blank">Slides</a>'
  ),
  Exercises = c(
    '<a href="analysis/QG3_Polygenic_Scores_practical.html">Practical</a>',
    ""
  ),
  Key = c("", "")
) %>%
kable(escape = FALSE, caption = "**Wednesday, June 10th**", align = c('l', 'c', 'l', 'c', 'c', 'c')) %>%
kable_styling(bootstrap_options = c("hover", "condensed", "striped"),
              full_width = FALSE,
              font_size = 14) %>%
column_spec(1, width = "12%") %>%
column_spec(2, width = "8%") %>%
column_spec(3, width = "35%")
```


## Datasets

All the data sets used can be downloaded from the [Box folder](https://gatech.box.com/s/9x413qjl5cprqmkvh9xt1f105tysxtcw) for the class.

## Additional resources

* [2016 TPB Paper](https://faculty.washington.edu/tathornt/sisg/Zheng_Weir_Eigenanalysis_TPB_2016.pdf) on PCA for Population Structure Inference and Admixture Estimation
* [HapMap MXL Individual Ancestry Figure](https://faculty.washington.edu/tathornt/sisg/HapMapMXL_Ancestry_Supervised.pdf) (Supervised Learning)
* If you are new to R, you may find the following material helpful:
  * [Introduction to R materials](https://faculty.washington.edu/kenrice/rintro/) from a past SISG Module
  * [Graphics with ggplot2](https://ggplot2.tidyverse.org/)
  * [Data manipulation with dplyr](http://dplyr.tidyverse.org/)
